Two applications of logic programming to Coq

### 2020-11-07 papers/types2020/elab.tex
Various minor updates, mostly to normalize the name of some FPC predicates.
small fixes before last pass to elab

To give an example, consider using untyped $\lambda$-terms encoded
using De Bruijn's notation as proof certificates for the propositional
intuitionistic logic over just $\supset$.  The fact that such terms
can be used as proof certificates for such formulas (denoting simple
types) can be formally defined using the FPC description in
Figure~\ref{fig:debruijn}.
%
Using the constants provided in that figure, the untyped
$\lambda$-term $\lambda x (x (\lambda y (y (\lambda z (x (\lambda
u\,z))))))$ can be encoded as the following Elpi term of type
\lsti{deb}.
%
\begin{lstlisting}[basicstyle=\ttfamily,language=lprolog]
(lambda (apply 0 [lambda (apply 0 [lambda (apply 2 
                                                    [lambda (apply 1 [])])])]))
\end{lstlisting}
Using the constructor \lsti{lc} and \lsti{args}, terms in De Bruijn
syntax (terms of type \lsti{deb}) are incorporated into proof
certificates (terms of type \lsti{cert}) along with other integer
arguments that are needed to compute offsets to address bound
variables.

>>>>

To give an example, consider using untyped $\lambda$-terms encoded
using De Bruijn's notation as proof certificates for the propositional
intuitionistic logic over just $\supset$.  The fact that such terms
can be used as proof certificates for such formulas (denoting simple
types) can be formally defined using the FPC description in
Figure~\ref{fig:debruijn}.
%
Using the constants provided in that figure, the untyped
$\lambda$-term $\lambda x (x (\lambda y (y (\lambda z (x (\lambda
u\,z))))))$ can be encoded as the following Elpi term of type
\lsti{deb}.
%
\begin{small}
\begin{lstlisting}[basicstyle=\ttfamily,language=lprolog]
(lambda (apply 0 [lambda (apply 0 [lambda (apply 2 [lambda (apply 1 [])])])]))
\end{lstlisting}
\end{small}
Using the constructor \lsti{lc} and \lsti{args}, terms in De Bruijn
syntax (terms of type \lsti{deb}) are incorporated into proof
certificates (terms of type \lsti{cert}) along with other integer
arguments that are needed to compute offsets to address bound
variables.

### 2020-11-07 papers/types2020/intro.tex
small fixes before last pass to elab

Recently, Enrico Tassi et al.\ developed the Elpi
implementation \cite{dunchev15lpar} of \lP \cite{miller12proghol} and
more recently, Tassi has made Elpi available as the Coq-Elpi
plugin~\cite{tassi18coqpl} (\url{https://github.com/LPCIC/coq-elpi})
to the Coq proof assistant.  This implementation of \lP extends
earlier ones in primarily two directions: First, Elpi adds a notion of
constraints and constraints handling rules, which makes it more
expressive than the Teyjus implementation \cite{nadathur99cade} of
\lP. Second, the plugin version of Elpi comes equipped with a
quotation and anti-quotation syntax for mixing Coq expressions with
\lP program elements and an API for accessing the Coq environment,
type checking included.

@@

Since relations (not functions) are central in \lP, Elpi is capable to
providing direct support for the many relations that have a role in
proof assistants implementation and usage.  Such relations include
typing (e.g., $\Gamma\vdash M\colon\sigma$), evaluation (e.g., natural
semantics specifications \cite{kahn87stacs,hannan93jfp}), and
interaction (e.g., structured operational semantics
\cite{milner89book,plotkin81}).

Felty has also made the point that LCF-style tactics and tacticals can
be given an elegant and natural specification using the higher-order
relational specifications provided by \lP \cite{felty93jar}.  Some
recent implementations built using Elpi support the usefulness of
higher-order logic programming as a meta-programming language for
proof assistants in general~\cite{DunchevCT16,coen19mscs} and, in
particular, for Coq via the Coq-Elpi plug~\cite{CohenST20,tassi19itp} .

In this paper, we present two applications of Elpi within Coq. With
these examples, we shall illustrate that Elpi is useful not only
because of its meta-programming features, but also because it soundly
implements a higher-order intuitionistic logic: such implementations
of higher-order logic have long been known to provide powerful and
flexible approaches to implementing many different logics and their
proof systems \cite{felty89phd,felty88cade,paulson89jar}.  Following
that tradition, the Elpi system makes it possible to encode the proofs
and proof theory of various subsets of the logic behind Coq (see also
\cite{felty93lics}).

>>>>

Recently, Enrico Tassi et al.\ developed the Elpi
implementation~\cite{dunchev15lpar} of \lP~\cite{miller12proghol} and
more recently, Tassi has made Elpi available as the Coq-Elpi
plugin~\cite{tassi18coqpl} (\url{https://github.com/LPCIC/coq-elpi})
to the Coq proof assistant.  This implementation of \lP extends
earlier ones in primarily two directions: First, Elpi adds a notion of
constraints and constraints handling rules, which makes it more
expressive than the Teyjus implementation~\cite{nadathur99cade} of
\lP. Second, the plugin version of Elpi comes equipped with a
quotation and anti-quotation syntax for mixing Coq expressions with
\lP program elements and an API for accessing the Coq environment,
type checking included.

@@

Since relations (not functions) are central in \lP, Elpi is capable to
providing direct support for the many relations that have a role in
proof assistants implementation and usage.  Such relations include
typing (e.g., $\Gamma\vdash M\colon\sigma$), evaluation (e.g., natural
semantics specifications~\cite{kahn87stacs,hannan93jfp}), and
interaction (e.g., structured operational semantics~\cite{milner89book,plotkin81}).

Felty has also made the point that LCF-style tactics and tacticals can
be given an elegant and natural specification using the higher-order
relational specifications provided by \lP~\cite{felty93jar}.  Some
recent implementations built using Elpi support the usefulness of
higher-order logic programming as a meta-programming language for
proof assistants in general~\cite{DunchevCT16,coen19mscs} and, in
particular, for Coq via the Coq-Elpi plug~\cite{CohenST20,tassi19itp}.

In this paper, we present two applications of Elpi within Coq. With
these examples, we shall illustrate that Elpi is useful not only
because of its meta-programming features, but also because it soundly
implements a higher-order intuitionistic logic: such implementations
of higher-order logic have long been known to provide powerful and
flexible approaches to implementing many different logics and their
proof systems~\cite{felty89phd,felty88cade,paulson89jar}.  Following
that tradition, the Elpi system makes it possible to encode the proofs
and proof theory of various subsets of the logic behind Coq (see also~\cite{felty93lics}).

### 2020-11-07 papers/types2020/pbt.tex
small fixes before last pass to elab

While nothing prevents us from porting all the PBT features we have
accounted for in~\cite{blanco19ppdp}, for the sake of this paper we
will implement only FPC corresponding to different flavors of
\emph{exhaustive} generation, as adopted, e.g., in
SmallCheck~\cite{smallcheck} and
$\alpha$Check~\cite{cheney_momigliano_2017}, and their
combination. Note however that it would take  no more than two lines of
code in the \lsti{decideE} expert to implement a form of random data
generation in the sense of randomized
backtracking~\cite{pltredexconstraintlogic}.

Of course, Coq already features \textsf{QuickChick}~\cite{QChick}
(\url{https://softwarefoundations.cis.upenn.edu/qc-current}), which is
a sophisticated and well-supported PBT tool, based on a different
perspective: being a clone of Haskell's QuickCheck, it emphasizes
testing executable (read \emph{decidable)} specifications with random
generators. While current research~\cite{LampropoulosPP18} aims to
increase automation, it is fair to say that testing with
\textsf{QuickChick}, in particular of relational specifications, is
still very labor intensive. We do not intend to compete with
\textsf{QuickChick} at this stage, but we shall see that we can test immediately
\lsti{Inductive} definitions that corresponds to pure Horn
programs. We can do that without having to provide a decidability
proof for those definitions. Furthermore, we are not committed to a
fixed random generation strategy, which in general requires additional
work in the configuration of generators and shrinkers.

@@

If we view  a Horn property (in uncurried form) as a many-sorted logical formula:
\[
  \forall x_1 \colon \tau_1\dots x_n \colon \tau_m, A_1\land\dots\land A_m \supset B\tag{*}\label{eq:prop}
  \]
%\(\forall x [(\tau(x)\wedge P(x)) \supset Q(x)]\) where $\tau$ is a
%typing predicate and
  where the $A_i$ and $B$ are  predicates defined
using Horn clause specifications, a counter-example to this conjecture
consists of a witness of the negated formula
\[
  \exists x_1 \colon \tau_1\dots x_n \colon \tau_n [A_1\land\dots\land A_m\wedge \neg B]
  \tag{**}\label{eq:full}
\]
%\(\exists x [(\tau(x)\land P(x)) \land \neg Q(x)]\).

In our Coq  setting $A_i$ and $B$ will be \lsti{Inductive}
propositions, while the $\tau_j$ are honest-to-goodness datatypes. We
will treat the two quite differently, in so far as elements of those
datatypes will be \emph{generated}, while predicates will only be
\emph{checked}. This distinction is reminiscent of bi-directional
type-checking and  plays also a part in interpreting the
negation sign.
%
When using a focused proof system for logic extended with fixed
points~\cite{baelde12tocl}, negation corresponds to the usual
intuitionistic interpretation, which is  what Coq supports.  % proofs of formulas such as
% \[
%   \exists x [(\tau(x)\wedge P(x)) \wedge \neg Q(x)]
%   \tag{*}\label{eq:full}
% \]
% are a single \emph{bipole}: that is, when reading a proof bottom up, a
% positive phase is followed on all its premises by a single negative
% phase that completes the proof.%
% In particular, the positive phase corresponds to the \emph{generation} phase
% and the negative phase corresponds to the \emph{testing} phase.
% %
%
% Instead of giving a full focused proof system of a logic including
% fixed points (since, as we will argue, that proof system will not, in
% fact, be needed to account for PBT ), we offer the following analogy.
% %
% Suppose that we are given a finite search tree and we are asked to
% prove that there is a secret located in one of the nodes of that
% tree.
% %
% A proof that we have found that secret can be taken to be a
% description of the path to that node from the root of the tree: that
% path can be seen as the proof certificate for that claim.
% %
% On the other hand, a proof that no node contains the secret is a
% rather different thing: here, one expects to use a blind and
% exhaustive search (via, say, depth-first or breath-first search) and
% that the result of that search never discovers the secret.
% %
% A proof of this fact requires \emph{no} external information: instead it
% requires a lot of computation involved with exploring the tree until
% exhaustion.
% %
% This follows the familiar pattern where the positive (generate) phase
% requires external information while the negative (testing) phase
% requires none.
%
However, for the sake of PBT and as we argued in~\cite{blanco19ppdp}  , we can identify a proof certificate for
$(\ref{eq:full})$ simply with a proof certificate for
$
 \exists x_1 \colon \tau_1\dots x_n \colon \tau_n [A_1\land\dots\land A_m]
$
and we can resort to negation-as-failure to check that the conclusion does not hold without caring for any evidence for the latter. This also means that we do not produce a Coq proof terms for the refutation of our property --- and neither does \textsf{QuickChick}, which runs at the OCaml level --- although we can return the witness for the existential.
% Such a certificate would contain the witness (closed) term $t$ for the
% existential quantifier and sufficient information to confirm that
% $P(t)$ can be proved (a proof of a typing judgment such as $\tau(t)$
% is usually trivial).
% Since a proof certificate for the existence-of-a-counterexample formula
% $(\ref{eq:full})$ can be taken as a proof certificate of
% $(\ref{eq:short})$, then we only need to consider proof certificates
% for Horn clause programs.
%
%Since the logic of Horn clauses is rather simple,
% We illustrate next what such proofs and proof certificates look like
% for the rather simple logic of Horn clauses.

\subsection{An Elpi tactic for PBT}
\label{ssec:dep_pbt}
We will invoke the tactic in a proof environment where the overall
goal is % one is trying to prove is the
the property that the system-under-test (SUT) should meet. This means that,
after \lsti|intro| has been used to introduce the relevant hypotheses,
% we have a proof
% context % consisting of typing information for a bunch of variables
% where eigenvariables play different roles: dependent variables, i.e., the ones appearing also in the types of other variables, are
% subject to \emph{generation}, while non-dependent ones tag assumptions that
% are only to tested.  % Obviously, the
% % two are the result of and \lsti|intro| on either a dependent or a
% % non-dependent product
% % ; in our setting this division acquires an additional reading beyond
% % what the property looks like: since specifications are encoded as pure
% % Horn clauses where no dependency is involved, non-dependent variables
% % simply tag the specification clauses, whereas dependent variables
% % represent the typing information for the data variables that the
% % specifications use.
% Within the tactic
the user 
specifies which variables of the environment should be used for
generating data and which for executing the specification. In addition
to this, the user should specify all the certificate information that
will guide the data generation phase. % The goal of the current
% environment will directly provide the property for which we want to
% find a counterexample.
% Say we want to generate a counterexample for
% some specification represented by \lsti|H1| and \lsti|H2| in the
% environment, and we want to generate the data for some variables
% \lsti|v1| and \lsti|v2| with a height bounded to 10: the final call to
% the tactic will then look like:
% \begin{lstlisting}
%   elpi pbt 10 (H1 /\ H2) (v1) (v2)
% \end{lstlisting}
Concretely, for the property~$(\ref{eq:prop})$
% \[
% \forall x_1 \colon \tau_1\dots x_n \colon \tau_m, A_1\land\dots\land A_m \supset B
% \]
and the specification of a FPC, the call to the tactic will be:
\begin{lstlisting}
elpi dep_pbt <fpc> $(A_1\land\dots\land A_m) \ (x_1) \dots (x_n)$.
\end{lstlisting}
The tactic calls \lsti{check} with the given FPC on the dependent
variables and will simply call a vanilla meta-interpreter (see
Fig.~\ref{fig:interp} in the appendix) to test the hypotheses and the negation of the
goal:
\begin{lstlisting}
interp $(A_1\land\dots\land A_m$), not (interp $B$)
\end{lstlisting}
where \lsti{not} is \lP's negation-as-failure operator.

To exemplify, let us add to the previous specification of list
insertion a definition of ordered list:
\begin{lstlisting}
Inductive ordered : list nat -> Prop :=
| o_n : ordered []
| o_s : forall x : nat, ordered [x]
| o_c : forall (x y : nat), forall xs,  ordered xs  -> x <= y -> ordered (x:: y :: xs).     
\end{lstlisting}
%
A property we may wish to check before embarking in a formal proof is
whether insertion preserves ordered-ness:
\begin{lstlisting}
Conjecture ins_ord:  forall (x : nat) xs rs, ordered xs -> insert x xs rs -> ordered rs.
intros x xs rs Ho Hi.
elpi dep_pbt height 5 (Ho /\ Hi ) (x) (xs).
Abort.
\end{lstlisting}
In this query the tactic tests the hypotheses \lsti{Ho} and \lsti{Hi}
against data \lsti{x,xs} generated exhaustively up to height
at most $5$  form the library \lsti{Inductive} definitions of
\lsti{nat} and \lsti{list}. We do not generate values for \lsti{rs},
since by mode information we know that it will be computed.
%
Since we did slip in an error (where?), our tactic reports a
counter-example, namely \verb|Proof Term: [0, [0; 1; 0]]|, and
we therefore abort the proof.

In order to generate the PBT query, some
pre-processing is needed. In particular, we need to turn the data
eigenvariables into \lP logic variables when they appear inside a
specification, and to generate queries %to \lsti|check|
for each of
these logic variables in association with the type of the data
variable it corresponds to.  In order to realize this pre-processing step,
we leverage extensively \lP's higher order programming features. The
substitutions are handled with the technique of \lsti|copy| clauses~\cite{miller91jlc}.

@@

For a more significant case study, let us turn to the semantics of
programming languages, where PBT has been used extensively and
successfully~\cite{Klein12}. Here we will consider a far simpler example,
a typed arithmetic language featuring numerals with predecessor  and test-for-zero, 
and Booleans with if-then-else, which comes from the \emph{Software Foundations} book series (\url{https://softwarefoundations.cis.upenn.edu/plf-current/Types.html}):
\begin{lstlisting}
Inductive tm : Type :=
 | ttrue : tm | tfalse : tm | tif : tm -> tm -> tm -> tm | tzero : tm | tsucc : tm -> tm
 | tpred : tm -> tm | tiszero : tm -> tm.
 Inductive typ : Type := |TBool : typ |TNat : typ.
\end{lstlisting}
The completely standard static and small step dynamic semantics are
reported in appendix~\ref{ssec:types}.

>>>>

While nothing prevents us from porting all the PBT features that we have
accounted for in~\cite{blanco19ppdp}, for the sake of this paper we
will implement only FPC corresponding to different flavors of
\emph{exhaustive} generation, as adopted, e.g., in
SmallCheck~\cite{smallcheck} and
$\alpha$Check~\cite{cheney_momigliano_2017}, and their
combination. Note however that it would take  no more than two lines of
code in the \lsti{decideE} expert to implement a form of random data
generation in the sense of randomized
backtracking~\cite{pltredexconstraintlogic}.

Of course, Coq already features \textsf{QuickChick}~\cite{QChick}
(\url{https://softwarefoundations.cis.upenn.edu/qc-current}), which is
a sophisticated and well-supported PBT tool, based on a different
perspective: being a clone of Haskell's QuickCheck, it emphasizes
testing executable (read \emph{decidable}) specifications with random
generators. While current research~\cite{LampropoulosPP18} aims to
increase automation, it is fair to say that testing with
\textsf{QuickChick}, in particular of relational specifications, is
still very labor intensive. We do not intend to compete with
\textsf{QuickChick} at this stage, but we shall see that we can test immediately
\lsti{Inductive} definitions that corresponds to pure Horn
programs, without having to provide a decidability
proof for those definitions. Furthermore, we are not committed to a
fixed random generation strategy, which in general requires additional
work in the configuration of generators and shrinkers.

@@

If we view  a Horn property (in uncurried form) as a many-sorted logical formula
\[
  \forall x_1 \colon \tau_1\dots x_n \colon \tau_m, A_1\land\dots\land A_m \supset B\tag{*}\label{eq:prop}
  \]
%\(\forall x [(\tau(x)\wedge P(x)) \supset Q(x)]\) where $\tau$ is a
%typing predicate and
  where the $A_i$ and $B$ are  predicates defined
using Horn clause specifications, a counter-example to this conjecture
consists of a witness of the negated formula
\[
  \exists x_1 \colon \tau_1\dots x_n \colon \tau_n , A_1\land\dots\land A_m\wedge \neg B
  \tag{**}\label{eq:full}
\]
%\(\exists x [(\tau(x)\land P(x)) \land \neg Q(x)]\).

In our Coq  setting $A_i$ and $B$ will be 
propositions, while the $\tau_j$ are honest-to-goodness datatypes. We
will treat the two quite differently, in so far as elements of those
datatypes will be \emph{generated}, while predicates will only be
\emph{checked}. This distinction is reminiscent of bi-directional
type-checking and  plays also a part in interpreting the
negation sign.
%
In a proof system for intuitionistic logic extended with fixed
points~\cite{baelde12tocl}, negation corresponds to the usual
intuitionistic interpretation, which is  what Coq supports.  % proofs of formulas such as
% \[
%   \exists x [(\tau(x)\wedge P(x)) \wedge \neg Q(x)]
%   \tag{*}\label{eq:full}
% \]
% are a single \emph{bipole}: that is, when reading a proof bottom up, a
% positive phase is followed on all its premises by a single negative
% phase that completes the proof.%
% In particular, the positive phase corresponds to the \emph{generation} phase
% and the negative phase corresponds to the \emph{testing} phase.
% %
%
% Instead of giving a full focused proof system of a logic including
% fixed points (since, as we will argue, that proof system will not, in
% fact, be needed to account for PBT ), we offer the following analogy.
% %
% Suppose that we are given a finite search tree and we are asked to
% prove that there is a secret located in one of the nodes of that
% tree.
% %
% A proof that we have found that secret can be taken to be a
% description of the path to that node from the root of the tree: that
% path can be seen as the proof certificate for that claim.
% %
% On the other hand, a proof that no node contains the secret is a
% rather different thing: here, one expects to use a blind and
% exhaustive search (via, say, depth-first or breath-first search) and
% that the result of that search never discovers the secret.
% %
% A proof of this fact requires \emph{no} external information: instead it
% requires a lot of computation involved with exploring the tree until
% exhaustion.
% %
% This follows the familiar pattern where the positive (generate) phase
% requires external information while the negative (testing) phase
% requires none.
%
However, for the sake of PBT and as we argued in~\cite{blanco19ppdp}  , we can identify a proof certificate for
$(\ref{eq:full})$ simply with a proof certificate for
$
 \exists x_1 \colon \tau_1\dots x_n \colon \tau_n , A_1\land\dots\land A_m
$
and we can resort to negation-as-failure to check that the conclusion does not hold without caring for any evidence for the latter. This also means that we do not produce a Coq proof terms for the refutation of our property --- and neither does \textsf{QuickChick}, which runs at the OCaml level --- although we can return the witness for the existential.
% Such a certificate would contain the witness (closed) term $t$ for the
% existential quantifier and sufficient information to confirm that
% $P(t)$ can be proved (a proof of a typing judgment such as $\tau(t)$
% is usually trivial).
% Since a proof certificate for the existence-of-a-counterexample formula
% $(\ref{eq:full})$ can be taken as a proof certificate of
% $(\ref{eq:short})$, then we only need to consider proof certificates
% for Horn clause programs.
%
%Since the logic of Horn clauses is rather simple,
% We illustrate next what such proofs and proof certificates look like
% for the rather simple logic of Horn clauses.

\subsection{An Elpi tactic for PBT}
\label{ssec:dep_pbt}
We will invoke the tactic in a proof environment where the overall
goal is % one is trying to prove is the
the property that the system-under-test (SUT) should meet. This means that,
after \lsti|intro| has been used to introduce the relevant hypotheses,
% we have a proof
% context % consisting of typing information for a bunch of variables
% where eigenvariables play different roles: dependent variables, i.e., the ones appearing also in the types of other variables, are
% subject to \emph{generation}, while non-dependent ones tag assumptions that
% are only to tested.  % Obviously, the
% % two are the result of and \lsti|intro| on either a dependent or a
% % non-dependent product
% % ; in our setting this division acquires an additional reading beyond
% % what the property looks like: since specifications are encoded as pure
% % Horn clauses where no dependency is involved, non-dependent variables
% % simply tag the specification clauses, whereas dependent variables
% % represent the typing information for the data variables that the
% % specifications use.
% Within the tactic
the user 
specifies which variables of the environment should be used for
generating data and which for executing the specification. In addition
to this, the user should specify all the certificate information that
will guide the data generation phase. % The goal of the current
% environment will directly provide the property for which we want to
% find a counterexample.
% Say we want to generate a counterexample for
% some specification represented by \lsti|H1| and \lsti|H2| in the
% environment, and we want to generate the data for some variables
% \lsti|v1| and \lsti|v2| with a height bounded to 10: the final call to
% the tactic will then look like:
% \begin{lstlisting}
%   elpi pbt 10 (H1 /\ H2) (v1) (v2)
% \end{lstlisting}
Concretely, for the property~$(\ref{eq:prop})$
% \[
% \forall x_1 \colon \tau_1\dots x_n \colon \tau_m, A_1\land\dots\land A_m \supset B
% \]
and the specification of a FPC, the call to the tactic will be:
\begin{lstlisting}
elpi dep_pbt <fpc> $(A_1\land\dots\land A_m) \ (x_1) \dots (x_n)$.
\end{lstlisting}
The tactic calls \lsti{check} with the given FPC on the dependent
variables and delegates to  a vanilla meta-interpreter (see
Fig.~\ref{fig:interp} in the appendix) the  test of the hypotheses and of the negation of the
goal:
\begin{lstlisting}
interp $(A_1\land\dots\land A_m$), not (interp $B$)
\end{lstlisting}
where \lsti{not} is \lP's negation-as-failure operator.

To exemplify, let us add to the previous specification of list
insertion a definition of ordered list:
\begin{lstlisting}
Inductive ordered : list nat -> Prop :=
| o_n : ordered []
| o_s : forall x : nat, ordered [x]
| o_c : forall (x y : nat), forall xs,  ordered xs  -> x <= y -> ordered (x:: y :: xs).     
\end{lstlisting}
%
A property we may wish to check before embarking in a formal proof is
whether insertion preserves ordered-ness:
\begin{lstlisting}
Conjecture ins_ord:  forall (x : nat) xs rs, ordered xs -> insert x xs rs -> ordered rs.
intros x xs rs Ho Hi.
elpi dep_pbt height 5 (Ho /\ Hi ) (x) (xs).
Abort.
\end{lstlisting}
In this query the tactic tests the hypotheses \lsti{Ho} and \lsti{Hi}
against data \lsti{x,xs} generated exhaustively up to height
at most $5$  from the library \lsti{Inductive} definitions of
\lsti{nat} and \lsti{list}. We do not generate values for \lsti{rs},
since by mode information we know that it will be computed.
%
Since we did slip in an error (where?), our tactic reports a
counter-example, namely \verb|Proof Term: [0, [0; 1; 0]]|, and
we therefore abort the proof.

In order to generate the PBT query, some
pre-processing is needed. In particular, we turn the data
eigenvariables into \lP logic variables when they appear inside a
specification, and  generate queries %to \lsti|check|
for each of
these logic variables in association with the type of the data
variable it corresponds to.  In order to realize this pre-processing step,
we leverage extensively \lP's higher order programming features. The
substitutions are handled with the technique of \lsti|copy| clauses~\cite{miller91jlc}.

@@

For a more significant case study, let us turn to the semantics of
programming languages, where PBT has been used extensively and
successfully~\cite{Klein12}. Here we will consider a far simpler example,
a typed arithmetic language featuring numerals with predecessor  and test-for-zero, 
and Booleans with if-then-else, which comes from the \emph{Software Foundations} book series (\url{https://softwarefoundations.cis.upenn.edu/plf-current/Types.html}):
\begin{lstlisting}
Inductive tm : Type :=
 | ttrue : tm | tfalse : tm | tif : tm -> tm -> tm -> tm | tzero : tm | tsucc : tm -> tm
 | tpred : tm -> tm | tiszero : tm -> tm.
 Inductive typ : Type := | TBool : typ | TNat : typ.
\end{lstlisting}
The completely standard static and small step dynamic semantics rules are
reported in appendix~\ref{ssec:types}.
